---
title: "Dashboard: Recently Archived CRAN Packages"

execute:
  freeze: false
---

<style>
tr { vertical-align: top; }
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  echo = FALSE
)

max_days <- 5*7
```

The below table lists all R packages that are no longer on CRAN,
because they have been either archived[^1] on or removed[^2] from CRAN
during the last `{r} max_days` days. The table shows the date when a
package was dropped by CRAN and their reason[^3] for it. If a package
has since been resubmitted and is waiting in the CRAN incoming queue,
it is marked as "resubmitted" (in the 'event' column).  Links to the
most recent CRAN checks and the source code are also provided.  If a
package does not return to CRAN within `{r} max_days` days, it is
removed also from CRANhaven[^4].

[^1]: When a package is _archived_, the archived version is still
available in the [CRAN archive area] together with all previous
versions.

[^2]: When a package is _removed_, which is a rare event, the removed
package version is completely removed from CRAN, and no longer
available for download. Earlier versions might be available in the
[CRAN archive area].

[^3]: The _reason_, which is based on the [CRAN Repository Policy], is
extracted from CRAN's manually curated [PACKAGES.in] database.

[^4]: If a package falls off also CRANhaven, it may still be
installable from the [CRAN mirror on R-universe].


```{r dashboard}
library("dplyr")

#' @importFrom jsonlite read_json
cranhaven_packages <- local({
  data <- NULL
  
  function(url = "https://raw.githubusercontent.com/cranhaven/cranhaven.r-universe.dev/main/packages.json") {
    if (!is.null(data)) return(data)
    db <- jsonlite::read_json(url)
    
    ## Patch missing fields
    names <- unique(unlist(lapply(db, FUN = names)))
    db <- lapply(db, FUN = function(x, names) {
      missing <- setdiff(names, names(x))
      for (name in missing) {
        x[[name]] <- NA_character_
      }
      as.data.frame(x)
    }, names = names)
    stopifnot(lengths(db) == length(db[[1]]))
    
    db <- do.call(rbind, db)

    ## Coerce to dates
    db$archived_on <- as.Date(db$archived_on)
    db$x_cran_comment_date <- as.Date(db$x_cran_comment_date)
    
    data <<- db
    data
  }
}) ## cranhaven_packages()

#' @importFrom ciw incoming
cran_incoming <- local({
  data <- NULL
  
  function() {
    if (!is.null(data)) return(data)
    db <- ciw::incoming(folder = c("pretest", "recheck", "inspect", "pending", "waiting", "newbies"), check = FALSE)
    db <- as.data.frame(db)
    colnames(db) <- tolower(colnames(db))
    db$package <- gsub("_.*", "", db$name)
    db$version <- gsub("(.*_|[.]tar[.]gz)", "", db$name)
    db <- db[, c("package", "version", "folder", "time")]
    data <<- db
    data
  }
})

captalize <- function(x) {
  vapply(x, FUN.VALUE = NA_character_, FUN = function(s) {
    if (is.na(s) || nchar(s) == 0) return(s)
    first <- substr(s, start = 1L, stop = 1L)
    tail <- substr(s, start = 2L, stop = nchar(s))
    paste0(toupper(first), tail)
  })
}

history <- cranhaven_packages()
incoming <- cran_incoming()
incoming <- incoming[, c("package", "folder")]
colnames(incoming)[2] <- "cran_incoming"
history <- merge(history, incoming, by = "package", all.x = TRUE)

## Rename columns
history <- rename(history,
  "date"   = archived_on,
  "event"  = x_cran_comment_event,
  "reason" = x_cran_comment_reason
)

idxs <- which(!is.na(history$cran_incoming))
history$event[idxs] <- paste(history$event[idxs], '<a href="https://r-hub.github.io/cransays/articles/dashboard.html">resubmitted</a>', sep = ", ")

history$reason <- gsub("'([[:alnum:].]+)'", '<a href="https://cran.r-project.org/package=\\1">\\1<a>', history$reason)

idxs <- which(!is.na(history$x_cran_history))
history$reason[idxs] <- paste(history$reason[idxs], sprintf("<br><small><em>Previously</em>: %s</small>", gsub("\n", " ", history$x_cran_history[idxs])), sep = ". ")
history$reason[-idxs] <- paste0(history$reason[-idxs], ".")
history$reason <- gsub("[.]+", ".", history$reason)
history$reason <- captalize(history$reason)

history$links <- with(history, paste(
  sprintf('<a href="https://cran-archive.r-project.org/web/checks/%s/%s_check_results_%s.html">checks</a>', sub("-.*", "", date), date, package),
  sprintf('<a href="https://github.com/cran/%s">source</a>', package),
  sep = ", "
))

history$package <- sprintf('<a href="https://cran.r-project.org/package=%s">%s</a>', history$package, history$package)

history <- history[, c("date", "package", "event", "reason", "links")]

# Show only recent packages
library("DT")
datatable(history,
  rownames = FALSE,
  escape = FALSE,
  elementId = "live-dashboard",
  options = list(
    pageLength = 50,
    order = list(list(0, "desc"), list(1, "asc"))
  )
)
```


[CRAN archive area]: https://cran.r-project.org/src/contrib/Archive/
[CRAN mirror on R-universe]: https://cran.r-universe.dev/builds
[CRAN Repository Policy]: https://cran.r-project.org/web/packages/policies.html
[PACKAGES.in]: https://cran.r-project.org/src/contrib/PACKAGES.in
